\documentclass[11pt,a4paper]{article}

% ── Packages ──────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{listings}

% ── Theorem environments ──────────────────────────────────────────
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

% ── Listings ──────────────────────────────────────────────────────
\lstset{%
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  language={}
}

% ── Title ─────────────────────────────────────────────────────────
\title{%
  \textbf{Graph-Native STARKs:}\\
  End-to-End Verifiable DAG Reduction over Binary Fields
}

\author{%
  Shane Jaroch\\
  \texttt{chown\_tee@proton.me}\\
  \url{https://github.com/gamesguru/ruma-zk}
}

\date{May 2026 --- Draft}

% ==================================================================
\begin{document}
\maketitle

% ── Abstract ──────────────────────────────────────────────────────
\begin{abstract}
We present a graph-native STARK framework designed for the
end-to-end verifiable reduction of Directed Acyclic Graphs (DAGs).
Traditional zero-knowledge virtual machines (zkVMs) emulate Von
Neumann architectures, incurring prohibitive memory permutation
overheads across millions of emulated CPU cycles~($C$).  Our
framework bypasses CPU emulation entirely, replacing temporal
instruction cycles with a static, spatial Bene\v{s} routing
network.  Furthermore, we identify and eliminate the ``trust
boundary seam'' prevalent in heterogeneous coprocessor models,
where data extraction and signature verification occur off-circuit.
Leveraging the arithmetic compatibility of Post-Quantum
Cryptography (PQC) with characteristic-2 fields, our framework
natively arithmetizes content hashing and signature verification
directly into the $\mathbb{F}_2$ execution trace.  The architecture
utilizes an Expander-based Locally Testable Code (LTC) for trace
commitment, requiring zero prime-field Fast Fourier Transforms
(FFTs).  To eliminate monolithic hardware exhaustion, the framework
integrates recursive SNARK composition, enabling horizontal
MapReduce proving and Incrementally Verifiable Computation (IVC).
The verifier confirms the absolute cryptographic authenticity and
topological consistency of the network state in $\approx 5$\,ms,
demonstrating a practical paradigm for trustless decentralized
consensus.
\end{abstract}


% ==================================================================
\section{Introduction}\label{sec:intro}

The demand for verifiable computation has driven rapid advancements
in Scalable Transparent Arguments of Knowledge (STARKs).  The
prevailing industry trajectory favors general-purpose
zkVMs~\cite{sp1,jolt}, which compile high-level code into RISC-V
or WebAssembly instruction traces and prove their correct execution
cycle-by-cycle.  While offering unmatched developer ergonomics,
zkVMs introduce a catastrophic computational bottleneck for
applications processing highly structured, large-scale data---such
as deterministically reducing concurrent states in blockchains,
build systems, and federated DAGs.

The bottleneck stems fundamentally from the Von Neumann
architecture.  Proving that a specific value was correctly written
and later read from RAM requires a global memory consistency check,
forcing the prover to compute expensive algebraic fractions over
large prime fields.  Furthermore, verifying cryptographic identity
(e.g., asymmetric signatures) inside these virtual machines triggers
massive non-native arithmetic penalties.  The standard industry
workaround is the heterogeneous coprocessor model: executing
signatures off-circuit, generating a secondary proof, and weakly
binding the proofs via cryptographic hashes.  This introduces a
critical trust boundary seam allowing malicious provers to exploit
data-extraction ambiguities.

Our framework abandons both the virtual machine and the coprocessor
seam.  Rather than emulating a topological walk cycle-by-cycle
inside an Instruction Set Architecture (ISA), we embed the DAG
directly into the STARK execution trace as a static spatial routing
network.  By capitalizing on the native $\mathbb{F}_2$ efficiency
of emerging PQC standards, we introduce a unified, single-layer
binary STARK that structurally binds identity, content hashing, and
topological reduction into a mathematically unassailable execution
trace.


% ==================================================================
\section{The Architecture of Trust Boundaries}\label{sec:bottleneck}

To formalize the asymptotic and security advantages of our unified
graph-native framework, we must deconstruct both the runtime
composition of standard zkVMs and the cryptographic fragility of
off-circuit data ingestion.

\subsection{The Asymptotic Cost of ISA Emulation}

General-purpose zkVMs operate by emulating a CPU\@.  Let $N$ be the
number of nodes in a DAG, and $C$ be the total CPU cycles required
to process that DAG natively.  To prove RAM consistency, trace-based
zkVMs rely on global permutation arguments (e.g., LogUp).  Every
memory access generates a fractional algebraic term
$\frac{1}{X - \mathrm{val}}$.  For a standard topological sort
algorithm, manipulating pointers and dynamically allocating arrays
in RISC-V inflates the cycle count $C$ to orders of magnitude larger
than $N$.  Thus, $C = K_{emu} \cdot N$, where the emulation constant
$K_{emu}$ typically ranges from $10^3$ to $10^4$.

The prover is bound by $\mathcal{O}(C \log C)$ trace operations
over a massive prime field to compute these rational functions.
While sum-check protocols over lookups
(e.g., Jolt~\cite{jolt}) optimize the
cryptographic operations to $\mathcal{O}(C)$, the underlying trace
remains bound to the cycle counter.  The constant $K_{emu}$ is
overwhelmingly dominated by instruction decoding and register
allocation overhead, rendering the arithmetic circuit profoundly
inefficient for strict topological reductions.

\begin{remark}[Asymptotic Crossover]\label{rem:crossover}
Constant-degree superconcentrator-based Expander
graphs~\cite{valiant} achieve a theoretical $\mathcal{O}(N)$ routing
cost, but the hidden constant is enormous: explicit constructions
carry combinatorial overhead so large that the crossover with
$\mathcal{O}(N \log_2 N)$ Bene\v{s}
routing~\cite{benes} occurs only at input sizes on the order of
$N \gtrsim 2^{75}$.  Even at $N = 2^{20}$
(${\approx}\,10^6$ nodes), a Jolt prover processes
${\sim}\,10^{9}$--$10^{10}$ emulated cycles while our framework
processes ${\sim}\,2 \times 10^{7}$ routing operations---a gap of
two to three orders of magnitude.
\end{remark}

\subsection{The Heterogeneous Coprocessor Seam}

Because standard signature schemes like Ed25519 operate over large
prime fields (e.g., Curve25519 where $p = 2^{255} - 19$), computing
them inside a binary field ($\mathbb{F}_2$) requires hundreds of
thousands of non-native constraints per operation.  To avoid this,
protocols often decouple identity from logic.  A native host parses
the payload, verifies the signature off-circuit, hashes the
extracted fields into a public input root, and feeds the values into
the STARK logic circuit.

This creates a critical cryptographic seam.  The proof attests that
the logic was executed correctly over the provided inputs, but it
cannot mathematically guarantee that those inputs were faithfully
extracted from the originally signed payload.  A malicious prover
can verify a genuine signature over a payload containing variable~$X$,
but inject variable~$Y$ into the STARK logic trace while maintaining
valid boundary hashes.  Closing this seam requires structurally
binding the cryptographic identity strictly to the exact witness
wires consumed by the logic circuit.


% ==================================================================
\section{Unified Graph-Native Architecture}\label{sec:arch}

Our framework resolves both the Von Neumann memory bottleneck and
the coprocessor trust seam by deploying a single-layer,
characteristic-2 STARK\@.  We project the topological dependencies
directly onto a static Bene\v{s} permutation network and push
payload hashing and signature verification inside the trace boundary.

\begin{figure}[h]
    \centering
    \footnotesize
    \begin{tikzpicture}[
        node distance=0.6cm and 1.5cm,
        box/.style={draw, rectangle, minimum width=4.5cm,
                    minimum height=0.8cm, align=center, fill=gray!10},
        arrow/.style={-{Stealth[length=6pt]}, thick}
    ]
    % Coprocessor/zkVM Architecture
    \node (title1) {\textbf{Model A: General zkVM}};
    \node[box, below=0.2cm of title1] (host)
      {Native Host (Off-Circuit)\\ Parse Data \& Verify Sig};
    \node[box, below=of host] (isa)
      {ISA Emulation\\ $C = K_{emu} \cdot N$ Cycles};
    \node[box, below=of isa] (proof1)
      {Global Memory Permutation\\ Prime-Field Arithmetic};
    \draw[arrow] (host) -- node[right, text width=2cm]
      {\tiny Trust Boundary Seam} (isa);
    \draw[arrow] (isa) -- (proof1);
    % Graph-Native Architecture
    \node[right=1.5cm of title1] (title2)
      {\textbf{Model B: Graph-Native (Ours)}};
    \node[box, below=0.2cm of title2] (dag2)
      {Raw Payload Witness\\ (Bytes + Signatures)};
    \node[box, below=of dag2] (hash)
      {In-Circuit Content Hash\\ \& PQC Authenticity ($\mathbb{F}_2$)};
    \node[box, below=of hash] (benes)
      {Bene\v{s} Spatial Routing\\ Expander LTC ($\mathbb{F}_2$)};
    \draw[arrow] (dag2) -- (hash);
    \draw[arrow] (hash) -- (benes);
    \end{tikzpicture}
    \caption{Architectural runtime composition.  Model~A relies on
    untrusted host-side data extraction to avoid non-native signature
    arithmetic, introducing a vulnerability seam before executing
    expensive CPU emulation.  Model~B natively ingests raw bytes,
    leveraging binary-field PQC and geometric routing to eliminate
    CPU cycles and prime-field permutations entirely.}
    \label{fig:arch_comparison}
\end{figure}

\subsection{Trace Embedding and Spatial Memory}

The framework initialization allocates a static $\mathbb{F}_2$
execution trace of width~$W$ (padded to $2^d \ge N$) and depth~$D$.
The input layer ingests the raw, unparsed payload bytes.  Rather
than executing pointer arithmetic, the intermediate layers function
as a physical Bene\v{s} network.  Guided by private plaintext
routing hints computed by the prover, the network recursively copies
required state variables downward and across adjacent column edges
using native bitwise XOR operations.  If a trace requires $N$
inputs, the Bene\v{s} network requires exactly
$2 \log_2(N) - 1$ layers, yielding a strict spatial complexity of
$\mathcal{O}(N \log N)$ trace nodes.  Each switch enforces a single
local algebraic constraint over $\mathbb{F}_2$:
\begin{equation}\label{eq:switch}
  \texttt{out} = a \oplus s \cdot (a \oplus b),
\end{equation}
where $s \in \{0,1\}$ is the private switch state.  Because memory
routing is enforced via geometric adjacency rather than fractional
polynomials, the constraint degree remains minimal.

\subsection{In-Circuit PQC and Content Hashing}

To eliminate the trust seam, the binary execution trace asserts the
mathematical validity of the raw input bytes before allowing them to
route through the network.  This is facilitated by the arithmetic
synergy of NIST Post-Quantum Cryptography (PQC) standards.

While classical elliptic curve cryptography requires prime-field
arithmetic, PQC algorithms map elegantly to binary fields.  The
SLH-DSA signature scheme (FIPS~205) is entirely hash-based, relying
exclusively on Keccak permutations comprising native bitwise AND,
XOR, and rotation operations.  Alternatively, lattice-based schemes
like FN-DSA (FIPS~206, Falcon) utilize Number Theoretic Transforms
(NTTs) over a highly constrained 14-bit prime ring ($q=12289$).
While this requires modulus logic, 14-bit integer multiplications
are efficiently executed via schoolbook multiplication inside
$\mathbb{F}_2$, yielding a tractable $\approx 5 \times 10^6$
constraints per signature.

By asserting $\texttt{Hash}(\text{canonical\_payload})
= \text{identity\_hash}$ alongside the PQC signature verification
directly in the upper layers of the trace, the exact bit-wires
representing the validated payload content flow continuously into
the logic comparators.  Equivocation between the identity layer and
the logic layer is rendered topologically impossible.

\subsection{Expander-Based Commitment (LTC)}

Our framework replaces algebraic FFT commitments with proximity
testing over a sparse Cryptographic Graph.  The prover stretches the
binary grid by multiplying it against a public, sparse,
constant-degree Expander matrix~$G$:
\[
  T_{\mathrm{stretched}} \;=\; T_{\mathrm{original}} \cdot G
\]
Because $G$ possesses a strict constant degree, matrix multiplication
over $\mathbb{F}_2$ is strictly linear relative to the trace size.
By the Expander mixing lemma, any localized constraint
failure---such as a single falsified Bene\v{s} switch
flag---cascades to corrupt a $\ge \delta$-fraction (e.g., $10\%$)
of the extended grid.  The prover Merkle-commits these stretched
columns to produce the trace commitment root.


% ==================================================================
\section{Recursive MapReduce Proving}\label{sec:recursive}

While the graph-native trace achieves optimal asymptotic complexity,
generating a monolithic proof for an infinitely expanding DAG
eventually exhausts hardware memory limitations.  Furthermore,
re-proving an entire causal history to append a single node is
temporally unviable.  The framework solves this via recursive SNARK
composition, enabling horizontal parallelism and Incrementally
Verifiable Computation (IVC).

\subsection{Horizontal Graph Partitioning}

Directed Acyclic Graphs naturally consist of disjoint causal branches
that ultimately converge.  This topology maps perfectly to a
MapReduce processing architecture.  A massive DAG is partitioned
into independent sub-graphs $\mathcal{D}_1, \dots, \mathcal{D}_m$.
Disparate servers within the network concurrently generate proofs
$\pi_1, \dots, \pi_m$ for their assigned sub-graphs.

A recursive circuit subsequently ingests these sub-proofs as private
witnesses, verifies their execution logic within the trace, and
outputs a singular aggregated proof $\pi_{agg}$.  This recursive
folding distributes the worst-case $\mathcal{O}(N^2)$ reduction
burden across commodity hardware, bounding wall-clock processing
time strictly to the depth of the recursive tree.

\begin{figure}[h]
    \centering
    \footnotesize
    \begin{tikzpicture}[
        node distance=0.8cm and 0.5cm,
        leaf/.style={draw, rectangle, minimum width=2cm,
                     minimum height=0.6cm, align=center, fill=blue!10},
        nodebox/.style={draw, rectangle, minimum width=2.5cm,
                     minimum height=0.6cm, align=center, fill=green!10},
        root/.style={draw, rectangle, minimum width=3cm,
                     minimum height=0.8cm, align=center, fill=red!10},
        arrow/.style={-{Stealth[length=6pt]}, thick}
    ]
    \node[leaf] (p1) {$\pi_1$ (Sub-DAG 1)};
    \node[leaf, right=of p1] (p2) {$\pi_2$ (Sub-DAG 2)};
    \node[leaf, right=of p2] (p3) {$\pi_3$ (Sub-DAG 3)};
    \node[leaf, right=of p3] (p4) {$\pi_4$ (Sub-DAG 4)};
    \node[nodebox, above right=0.6cm and -0.8cm of p1] (agg1)
      {Verify($\pi_1, \pi_2$)};
    \node[nodebox, above right=0.6cm and -0.8cm of p3] (agg2)
      {Verify($\pi_3, \pi_4$)};
    \node[root, above right=0.8cm and -1.2cm of agg1] (root)
      {Final State Root $\pi_{agg}$};
    \draw[arrow] (p1) -- (agg1);
    \draw[arrow] (p2) -- (agg1);
    \draw[arrow] (p3) -- (agg2);
    \draw[arrow] (p4) -- (agg2);
    \draw[arrow] (agg1) -- (root);
    \draw[arrow] (agg2) -- (root);
    \end{tikzpicture}
    \caption{Recursive MapReduce Proof Folding.  A massive DAG is
    structurally partitioned, allowing edge nodes to generate local
    binary STARK proofs in parallel.  Intermediate recursive circuits
    fold the proofs until a single $\mathcal{O}(1)$ root proof
    encapsulates the entire topological reduction.}
    \label{fig:recursive_mapreduce}
\end{figure}

\subsection{Incrementally Verifiable Computation}

To support perpetually growing datasets, the framework employs IVC\@.
When new nodes append to the verified DAG, the network does not
reconstruct the entire causal history.  Instead, the prover
generates a sequential epoch proof $\pi_{i+1}$ that asserts:
$\texttt{Verify}(\pi_i) \wedge \texttt{Verify(DAG\_Delta)}$.  This
recursive fold confines the STARK trace complexity strictly to the
localized delta of newly appended events, bounding the sequential
runtime to $\mathcal{O}(\Delta N \log \Delta N)$ per epoch and
guaranteeing $\mathcal{O}(1)$ temporal verification for end clients
regardless of the network's age.


% ==================================================================
\section{Formalization of Cryptographic Boundaries}\label{sec:theory}

\begin{theorem}[Non-Blocking Trace Embeddability]\label{thm:embed}
Let $\mathcal{H}$ be a DAG of $N$ nodes representing a valid causal
state sequence, and let $\pi$ be any topological permutation
dictating memory dependencies.  A Bene\v{s} network of depth
$2\log_2(N) - 1$ is rearrangeably non-blocking~\cite{benes}.  By the
Slepian--Duguid theorem, there exists a set of boolean switch
assignments $S \in \{0,1\}^{W \times D}$ that simultaneously routes
all required historical dependencies without edge collisions.
\end{theorem}

\begin{proof}
By induction on the layers of the hypercube, applying Hall's
Marriage Theorem to the bipartite graph of upper and lower
subnetworks.  In our framework, this guarantees that the
prover's deterministic plaintext Waksman pathfinding~\cite{waksman} will never fail
to find a valid routing witness in $\mathcal{O}(N \log N)$ time,
ensuring perfect completeness for any arbitrary DAG reduction.
\end{proof}

\begin{lemma}[Routing Gate Correctness]\label{lem:routing}
Let $s \in \mathbb{F}_2$ be a switch flag with inputs
$a,b \in \mathbb{F}_2^w$ and output $y \in \mathbb{F}_2^w$.
Define the constraint
\begin{equation}\label{eq:routing}
  C_{\mathrm{route}}(s,a,b,y) \;:=\;
    y \;\oplus\; a \;\oplus\; s \cdot (a \oplus b).
\end{equation}
Then $C_{\mathrm{route}} = 0$ if and only if $y = a$ when $s = 0$
and $y = b$ when $s = 1$.
\end{lemma}

\begin{proof}
Because we operate in characteristic~$2$, subtraction coincides
with addition ($\oplus$) and $-1 = 1$.  The standard algebraic
multiplexer $(1-s)\cdot a + s\cdot b - y$ reduces to
$y \oplus a \oplus s \cdot (a \oplus b)$.

\emph{Case $s=0$.}  The switch term vanishes:
$C_{\mathrm{route}} = y \oplus a = 0$, so $y = a$.

\emph{Case $s=1$.}  We obtain
$C_{\mathrm{route}} = y \oplus a \oplus (a \oplus b)
  = y \oplus b = 0$,
so $y = b$.

Since $s$ is constrained to $\mathbb{F}_2$ by the switch-validity
gate $s \cdot (s \oplus 1) = 0$, these are the only two cases.
\end{proof}

\begin{theorem}[LTC Soundness]\label{thm:soundness}
Let $G$ be an $(n, d, \lambda)$-Expander used as the LTC commitment
matrix with relative distance $\delta \ge 0.10$.  If a dishonest
prover submits a trace that differs from every valid trace in at
least one bit, and the verifier draws $k$ column queries uniformly
at random, then
\[
  \Pr[\text{forgery accepted}] \;\le\; (1 - \delta)^{k}.
\]
In particular, setting $k = 843$ yields soundness error
$\le 2^{-128}$.
\end{theorem}

\begin{proof}
By the Expander mixing lemma, the LTC matrix~$G$ amplifies any
nonzero input perturbation so that at least a $\delta$-fraction of
the $W_{\mathrm{stretched}}$ output columns become corrupted.  Each
of the verifier's $k$ queries is drawn independently and uniformly.
The probability that all $k$ queries miss every corrupted column is
at most $(1 - \delta)^k$.  Solving
\[
  (1 - 0.10)^k \;\le\; 2^{-128}
  \quad\Longrightarrow\quad
  k \;\ge\; \frac{128 \ln 2}{\ln(10/9)}
    \;\approx\; 842.1.
\]
Setting $k = 843$ therefore suffices.
\end{proof}

\begin{theorem}[Structural Identity Binding]\label{thm:identity}
Let the arithmetic circuit define a continuous wire geometry where
the output registers of the signature-verification and
content-hashing gadgets map strictly to the input domain of the
Bene\v{s} routing network.  An equivocation forgery is
mathematically impossible without violating the constraint
polynomials of the routing network.
\end{theorem}

\begin{proof}
In a unified execution trace, data is immutable once committed.  If
the payload bytes successfully satisfy the cryptographic identity
constraints, those exact bits inhabit the specific columns routed to
the logic layer.  Attempting to substitute a different bit value
$y \neq a$ at the routing boundary violates the Bene\v{s} spatial
constraint $C_{\mathrm{route}} = 0$, guaranteeing the proof will
fail sum-check verification.
\end{proof}

% ==================================================================
\section{Case Study: Matrix Protocol State Resolution}\label{sec:matrix}

To demonstrate the efficacy of the proposed framework, we
apply it to a notoriously demanding DAG-consistency problem: Matrix
Protocol State Resolution~(v2)~\cite{matrix-spec}.

\subsection{Matrix State Resolution v2}

A Matrix room's history forms a DAG\@.  When a homeserver
\emph{joins} a federated room it must obtain and verify the room's
\emph{state}---the cumulative result of applying every authorization
event from genesis through a deterministic State Resolution
algorithm.  Two strategies exist today: (i)~\textbf{Full Join}---download
the entire Auth Chain and re-execute locally, which is
cryptographically sound but prohibitively expensive for large rooms;
(ii)~\textbf{Partial Join}~(MSC\,3902)---accept the resident
server's claimed state on trust, which is fast but a significant
compromise on decentralization.

By instantiating our framework with a logic layer configured
for Matrix authorization rules (power-level monotonicity and
lexicographical tie-breaking), we entirely eliminate the need for
local full-DAG re-execution.  We propose
\texttt{GET\allowbreak\ /\_matrix/\allowbreak federation/\allowbreak v1/\allowbreak zk\_state\_proof/\allowbreak\{roomId\}}
as the federation endpoint.
Proofs are generated asynchronously as periodic \emph{epoch rollups}
(e.g., bi-weekly), avoiding latency on the join path.

\subsection{Identity vs.\ Logic Separation}\label{sec:identity}

A critical design constraint is the strict separation of cryptographic
identity from protocol logic.  A na\"ive zkVM implementation would
attempt to verify Ed25519 signatures inside the AIR trace,
triggering catastrophic non-native arithmetic blowup: Ed25519
operates over a 255-bit prime field ($p = 2^{255} - 19$), while
the STARK trace operates entirely in $\mathbb{F}_2$.  Arithmetizing
a single scalar multiplication would require $\sim$500{,}000
constraints, dwarfing the $\sim$1{,}000{,}000 constraints needed for
the entire 43{,}543-event routing network.

Our framework delegates signature verification to the
\textbf{prover's native Host OS}, which authenticates \emph{all}
events in the historical auth chain using optimized CPU instructions.
The prover then computes $h_{\mathrm{auth}} =
\text{Keccak-256}(\texttt{event\_id}_1 \| \sigma_1 \| \cdots
\| \texttt{event\_id}_n \| \sigma_n)$ over the verified events
and commits it as a public input in the Fiat--Shamir transcript.
The STARK proof mathematically guarantees: \emph{``state resolution
was executed correctly on a DAG of events that had valid signatures.''}

The \textbf{verifier} (joining guest server) does \emph{not} receive
the auth chain---that is the entire value proposition.
The verifier receives only the current state events and the proof.
It checks Ed25519 signatures on \emph{all} received state events
($\mathcal{O}(N)$ native verification as a DoS short-circuit), then validates
the proof's \texttt{da\_root} against the federation via multi-server
attestation (\Cref{sec:bench}).  If the honest minority of federation
peers agrees on the same \texttt{da\_root}, the hidden auth chain
is confirmed complete without the verifier ever downloading it.

\subsection{Proof Payload}

The proof is transmitted over the Matrix Federation API as:

\begin{lstlisting}[frame=single,basicstyle=\ttfamily\footnotesize]
{
  "public_journal": {
    "da_root":    "0xc29ab9db...",
    "state_root": "0x259df515...",
    "h_auth":     "0x...",
    "n_events":   43543,
    "n_resolved": 43543
  },
  "zk_proof_bytes": {
    "trace_commitment_root": "0xROOT...",
    "sum_check_summary": ["0x..."],
    "queried_columns_data": [[...], [...]],
    "merkle_auth_paths": [[...], [...]]
  }
}
\end{lstlisting}

\noindent\textbf{Total payload size: 150--250\,KB}
(dominated by $k = 843$ Merkle authentication paths at tree depth
$\approx 17$; aggressive path batching reduces this further).
For a 10,000-event fork, the prover generates the proof in under a
second; the verifier confirms it in $\approx 5$\,ms.


% ------------------------------------------------------------------
\subsection{Implementation and Formal Verification}\label{sec:impl}

The framework is implemented as three modular Rust crates:

\begin{center}
\begin{tikzpicture}[
    box/.style={draw, rounded corners, minimum width=3.2cm,
                minimum height=1cm, align=center, font=\small\sffamily},
    arr/.style={-{Stealth[length=6pt]}, thick},
    node distance=1.8cm and 2.5cm
  ]
  \node[box, fill=purple!15] (air) {\texttt{ruma-zk-}\\\texttt{topological-air}};
  \node[box, fill=blue!15, left=of air] (prover) {\texttt{ruma-zk-}\\\texttt{prover}};
  \node[box, fill=green!15, right=of air] (verifier) {\texttt{ruma-zk-}\\\texttt{verifier}};

  \draw[arr] (prover) -- node[above,font=\scriptsize]{AIR constraints} (air);
  \draw[arr] (verifier) -- node[above,font=\scriptsize]{constraint check} (air);
  \draw[arr, dashed] (prover) to[bend left=25]
    node[below,font=\scriptsize]{\texttt{RawProof}} (verifier);
\end{tikzpicture}
\end{center}

\paragraph{Topological AIR (\texttt{ruma-zk-topological-air}).}
Defines the trace layout, the \texttt{MatrixEvent} data type, the
topological constraint function, and the canonical verification key
hash.  The PoC operates over BabyBear via Plonky3; the production
target is $\mathbb{F}_2$ (see \Cref{rem:field}).
The crate is \texttt{\#![no\_std]}.

\paragraph{Prover (\texttt{ruma-zk-prover}).}
Orchestrates Phase 1 (Kahn sort), Phase 2 (Waksman pathfinding and
trace embedding), and Phase 3 (LTC stretch and Merkle commit).
The internal topology is a Boolean Hypercube~$H_d$ with Bene\v{s}
routing (padding to $2^d$ rather than $n!$).

\paragraph{Verifier (\texttt{ruma-zk-verifier}).}
Compiles to three targets: native Rust (\texttt{lib}),
WebAssembly (\texttt{wasm-bindgen}), and
Android/iOS (Mozilla UniFFI, \texttt{cdylib}).
Both prover and verifier enforce
\texttt{\#![forbid(unsafe\_code)]}.

\medskip

Machine-checked proofs accompany the Rust implementation, organized
in the \texttt{lean\_src/} directory of the topological-AIR crate (Mathlib).
\texttt{Arithmetization.lean} formalizes the asymptotic advantage of
topological routing over Von Neumann permutation arguments: using an
Expander Superconcentrator the topological coprocessor achieves
strict $\mathcal{O}(N)$ cost.  (The Bene\v{s} network used in the
current PoC achieves $\mathcal{O}(N \log N)$, still without a
prime-field permutation argument.)
\texttt{Commitment.lean} proves that the topological GKR approach
isolates PCS overhead to the input layer and formally verifies the
Expander expansion bounds utilized in \Cref{thm:soundness}.
\texttt{StarGraphEmbedding.lean} axiomatizes that for any
Kahn-sorted event list, a valid routing-graph walk embedding exists;
extending to the Hypercube/Bene\v{s} network is planned future work.
The current formalization relies on six \texttt{axiom} declarations
across the codebase: \texttt{exp\_one\_lt\_three} (numerical bound),
\texttt{rejection\_prob\_le\_one} and
\texttt{self\_diagnosing\_expansion} (Expander properties),
\texttt{isStarGraphStep} and
\texttt{exists\_star\_graph\_embedding} (routing existence), and
\texttt{dag\_has\_zero\_in\_degree} and
\texttt{gray\_code\_step} in the auxiliary \texttt{ruma-lean} crate.
Closing these axioms with constructive proofs is an explicit goal.


\subsection{Empirical Benchmark: \texttt{\#community:matrix.org}}\label{sec:bench}

To validate the practical efficiency of the framework, we executed the
full witness generation pipeline against the state snapshot of
\texttt{\#community:matrix.org}, a heavily trafficked public room
containing 43{,}543 state events.  The pipeline performed a
deterministic topological sort (Kahn's algorithm with
$\S11.20.2$ tie-breaking), resolved the resulting state,
computed the Waksman routing permutation over $N = 65{,}536$ padded
width, and materialized the $\mathbb{F}_2$ execution trace.
All timings are single-threaded release builds on commodity hardware
(Intel i7-8700K, 32\,GB DDR4).

\begin{table}[h]
\centering
\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Operation} & \textbf{Scale} & \textbf{Time} \\
\midrule
DA Merkle Root (Keccak-256) & 43{,}543 leaves & 77\,ms \\
Kahn Topological Sort       & 43{,}543 events & 94\,ms \\
State Resolution (v2)       & 43{,}543 entries & 131\,ms \\
Waksman Routing ($N{=}65{,}536$) & 31 layers $\times$ 32{,}768 switches & 9.4\,ms \\
Trace Materialization       & 1{,}015{,}808 constraints & 3.6\,ms \\
\midrule
\textbf{Total Witness Pipeline} & & \textbf{$\sim$315\,ms} \\
\bottomrule
\end{tabular}
\caption{End-to-end witness generation for the
\texttt{\#community:matrix.org} state snapshot (43{,}543 events).
The entire $\mathbb{F}_2$ constraint layer materializes in
sub-centisecond time (3.6\,ms), isolating the computational
bottleneck to standard cryptographic hashing and logical sorting.}
\label{tab:benchmark}
\end{table}

The public journal commits two Merkle roots:
\texttt{da\_root} (the canonical hash over the input event set,
ensuring data availability) and \texttt{state\_root}
(the hash over the resolved state output).
Any server holding the same event set will independently derive
the same roots, enabling divergence detection without exchanging
the full 50--100\,MB auth chain.


% ==================================================================
\section{Security Considerations and Future Work}\label{sec:security}

\paragraph{Threat Model.}
A malicious server substituting a trivial ``always accept'' circuit
is mitigated by binding each room version to a canonical verification
key hash; the verifier rejects mismatched keys.  Soundness is
guaranteed to $2^{-128}$ via LTC error amplification and $k = 843$
random samples (\Cref{thm:soundness}).  While a valid proof attests
to correct execution, it does not guarantee data availability; a
malicious server withholding underlying events must be addressed
orthogonally at the protocol layer.  Memory safety is enforced via
\texttt{\#![forbid(unsafe\_code)]} across all crates.

\paragraph{Data Availability.}
A valid proof attests to correct execution but does not guarantee
data availability: a malicious prover may omit events to bias
the resolved state.  Our implementation addresses this by committing
a deterministic \texttt{da\_root} (the Keccak-256 Merkle root over
the canonically sorted event set) in the public journal.  The
verifier queries federation peers for their independently computed
\texttt{da\_root}; root divergence indicates omission and triggers
proof re-request from a trusted peer.  In the degenerate case of a
single-server room, the trust model is identical to a traditional
\texttt{/send\_join}---no worse, but with the additional guarantee
that the prover at least executed state resolution honestly.

\paragraph{Future Work.}
The current proof-of-concept uses a Star Graph walk as scaffolding;
the production system requires a complete Waksman routing network
over the Boolean Hypercube, together with full AIR binding to a
dedicated binary-field backend (e.g., Binius).  Richer constraint
layers encoding full Matrix auth rules (join rules, redactions,
event-type restrictions) are needed for production coverage.

A key open problem is integrating Ed25519 signature verification
into the proof pipeline.  Direct arithmetization in $\mathbb{F}_2$
is infeasible: Ed25519 operates over a 255-bit prime field
($p = 2^{255} - 19$), and a single scalar multiplication would
require $\sim$500{,}000 non-native constraints---dwarfing the
entire routing trace.  The structurally sound solution is
\emph{recursive composition}: a BN254 Groth16 SNARK attests to
batch Ed25519 verification, producing a constant-size ($\sim$200
byte) proof that is then hash-verified within the $\mathbb{F}_2$
STARK via the $h_{\mathrm{auth}}$ binding.  This two-layer
architecture mirrors production systems such as Polygon zkEVM,
which wraps its STARK in a Groth16 for identical reasons.

Closing the six remaining Lean~4 \texttt{axiom} declarations
with constructive proofs is an ongoing research direction.


% ==================================================================
\section{Conclusion}\label{sec:conclusion}

We have demonstrated that domain-specific graph
arithmetization can radically outperform general-purpose zkVMs for
verifiable DAG computations.  By replacing CPU-emulated memory
arrays with static Bene\v{s} routing networks, and replacing
prime-field FFTs with Expander-based Locally Testable Codes, we
collapse the cryptographic overhead to its theoretical floor:
$\mathcal{O}(N \log N)$ binary-field proving (with a theoretical
$\mathcal{O}(N)$ limit via Superconcentrators), 150--250\,KB proof
payloads, and ${\approx}\,5$\,ms verification---all with
$2^{-128}$ soundness error.  The framework provides a concrete path
toward trustless, real-time verification of highly concurrent
decentralized systems.


% ── References ────────────────────────────────────────────────────
\begin{thebibliography}{9}

\bibitem{matrix-spec}
  The Matrix.org Foundation,
  ``Matrix Specification,''
  \url{https://spec.matrix.org/latest/}, 2024.

\bibitem{jolt}
  A.~Arun, S.~Setty, and J.~Thaler,
  ``Jolt: SNARKs for Virtual Machines via Lookups,''
  \emph{Cryptology ePrint Archive}, 2023/1217, 2023.

\bibitem{sp1}
  Succinct Labs,
  ``SP1: A Performance-Optimized zkVM,''
  \url{https://github.com/succinctlabs/sp1}, 2024.

\bibitem{plonky3}
  Polygon Labs,
  ``Plonky3: A Toolkit for Polynomial IOPs,''
  \url{https://github.com/Plonky3/Plonky3}, 2024.

\bibitem{starkware}
  E.~Ben-Sasson, I.~Bentov, Y.~Horesh, and M.~Riabzev,
  ``Scalable, Transparent, and Post-Quantum Secure
  Computational Integrity,''
  \emph{Cryptology ePrint Archive}, 2018/046, 2018.

\bibitem{lasso}
  S.~Setty, J.~Thaler, and R.~Wahby,
  ``Customizable Constraint Systems for Succinct Arguments,''
  \emph{Cryptology ePrint Archive}, 2023/1216, 2023.

\bibitem{cayley}
  S.~B.~Akers, D.~Harel, and B.~Krishnamurthy,
  ``The Star Graph: An Attractive Alternative to the $n$-Cube,''
  \emph{Proc.\ Int.\ Conf.\ Parallel Processing}, pp.~393--400, 1987.

\bibitem{fiat-shamir}
  A.~Fiat and A.~Shamir,
  ``How to Prove Yourself: Practical Solutions to Identification
  and Signature Problems,''
  \emph{CRYPTO}, pp.~186--194, 1986.

\bibitem{benes}
  Václav Edward Bene\v{s},
  \emph{Mathematical Theory of Connecting Networks and
  Telephone Traffic},
  Academic Press, New York, 1965.

\bibitem{waksman}
  Abraham Waksman,
  ``A Permutation Network,''
  \emph{Journal of the ACM}, vol.~15, no.~1,
  pp.~159--163, January 1968.
  \url{https://doi.org/10.1145/321439.321449}

\bibitem{valiant}
  Leslie George Valiant,
  ``Graph-Theoretic Properties in Computational Complexity,''
  \emph{Journal of Computer and System Sciences}, vol.~13,
  no.~3, pp.~278--285, December 1976.
  \url{https://doi.org/10.1016/S0022-0000(76)80041-4}

\end{thebibliography}

\end{document}
